Cover Times of Random Walks

### 2019-04-15 paper.tex
matthews lower bound

\DeclareMathOperator{\E}{\mathrm{E}}		     % expected value
\DeclareMathOperator{\pr}{\mathrm{P}}		     % expected value
\DeclareMathOperator{\cov}{\uptau_\textrm{cov}}  % cover time variable
\DeclareMathOperator{\tcov}{t_\textrm{cov}}      % cover time
\DeclareMathOperator{\hit}{t_{\textrm{hit}}}     % hitting time
\DeclareMathOperator{\Ceff}{C_\textrm{eff}}      % effective conductance
\DeclareMathOperator{\Reff}{R_{\textrm{eff}}}     % effective resistance

@@

\section{Bounds}
\subsection{Matthews Method}
We have seen that it is in general very hard
to calculate the cover time of a random walk
on a graph.
The exceptions are symmetric or small graphs.
In lieu of an exact solution, we want to 
bound the cover time.

The Matthews Method is an example of a remarkably good bound \cite{LP17}.
In fact, it is tight for the complete graph:
the expected time between any two vertices
is $n-1 = \hit$ and \cref{thm:matthews} gives
us the main result of \cref{sec:complete}.

@@

\begin{theorem}
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews}
\end{theorem}
\begin{align}
\tcov &\leq \hit \left(1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{n-1} \right). \nonumber
\end{align}

@@

Finally, consider the expected time between $T_{k-1}$ and $T_{k}$
given that last vertex we visit is $\sigma_k$.
Certainly there exist $r$ and $s$ where $1 \leq r \neq s \leq n-1$
such that $L_{k-1} = r$ and $L_k = s$.
Then $\E_n[T_k - T_{k-1} | L(k) = \sigma_k] = \E_r[T_s] \leq \hit$.
So
\begin{align}
\E_n[T_k - T_{k-1}] &\leq
\hit \frac{1}{k}. \nonumber 
\end{align}

Putting it all together,
\begin{align}
\tcov \leq \hit \left( 1+ \frac{1}{2} + ... + \frac{1}{k} 
+ ... + \frac{1}{n-1} \right). \nonumber
\end{align}

\end{proof}

\subsection{Commute Time Identity}
Recall \cref{sec:electric}.
Define the probability $\pr_a(T_b < T_a)$ that starting at $a$
we reach $b$ before returning to $a$.
(Recall that $T_a>0$.)
Now define the effective conductance $\Ceff$ as $i_a/v_a$
and the effective resistance $\Reff$ as $1/\Ceff$.

>>>>

\DeclareMathOperator{\E}{\mathrm{E}}		     % expected value
\DeclareMathOperator{\pr}{\mathrm{P}}		     % expected value
\DeclareMathOperator{\cov}{\uptau_\textrm{cov}}  % cover time variable
\DeclareMathOperator{\tcov}{t_\textrm{cov}}      % cover time
\DeclareMathOperator{\hit}{t_{\textrm{hit}}}     % hitting time
\DeclareMathOperator{\tmina}{t_\textrm{min}^A}   % minimum time on A
\DeclareMathOperator{\Ceff}{C_\textrm{eff}}      % effective conductance
\DeclareMathOperator{\Reff}{R_{\textrm{eff}}}    % effective resistance

@@

\section{Bounds}

\subsection{Commute Time Identity}
We use the electrical network analysis in \cref{sec:electric} to 
prove the commute time identity so we may use it to help us bound
cover times.

Define the probability $\pr_a(T_b < T_a)$ that starting at $a$
we reach $b$ before returning to $a$.
(Recall that $T_a>0$.)
Now define the effective conductance $\Ceff$ as $i_a/v_a$
and the effective resistance $\Reff$ as $1/\Ceff$.

\subsection{Matthews Method}
We have seen that it is in general very hard
to calculate the cover time of a random walk
on a graph.
The exceptions are symmetric or small graphs.
In lieu of an exact solution, we want to 
bound the cover time.

The Matthews Method is an example of a remarkably good bound \cite{LP17}.
In fact, it is tight for the complete graph:
the expected time between any two vertices
is $n-1 = \hit$ and \cref{thm:matthews_up} gives
us the main result of \cref{sec:complete}.

@@

\begin{theorem}[Matthews Upper Bound]
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews_up}
\end{theorem}
\begin{align}
\tcov &\leq \hit \left(1 + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{n-1} \right). \nonumber
\end{align}

@@

Finally, we find the expected time between $T_{k-1}$ and $T_{k}$
given that last vertex we visit is $\sigma_k$.
Consider $r$ and $s$ where $1 \leq r \neq s \leq n-1$
such that $L_{k-1} = r$ and $L_k = s$.
Then $\E_n[T_k - T_{k-1} | L(k) = \sigma_k] = \E_r[T_s] \leq \hit$.
So
\begin{align}
\E_n[T_k - T_{k-1}] &\leq
\hit \frac{1}{k}. \nonumber 
\end{align}

Putting it all together,
\begin{align}
\tcov \leq \hit \left( 1+ \frac{1}{2} + ... + \frac{1}{k} 
+ ... + \frac{1}{n-1} \right). \nonumber
\end{align}
\end{proof}

The same strategy for the Matthews upper bound can be used
to find a lower bound.
Instead of looking for all $n$ vertices, we search
for $A \subseteq V(G)$.
When the hitting time is large between any two vertices in $A$,
the time to visit all of $A$ is a good lower bound on the cover time
of $V(G)$.

\begin{theorem}[Matthews Lower Bound]
Let $A \subseteq V(G)$. Define $\tmina = \min_{a,b\in A, a \neq b} \E_b[T_a]$.
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews_low}
\end{theorem}
\begin{align}
\tcov &\geq \tmina \left(1 + \frac{1}{2} + \frac{1}{3} + ... +
\frac{1}{|A| -1} \right). \nonumber
\end{align}

\begin{proof}
Without loss of generality, label the vertices $\{1, 2, ..., n\}$.
We may assume that the walk started at vertex $x \in A$.
Let $\sigma$ be a uniform permutation of the vertices in $A$.
The strategy is to look for the vertices in order $\sigma$.

Let $T_k$ be the first time that vertices $\sigma_1, \sigma_2, ..., \sigma_k$
have all been visited.
For ease of notation, set $L_k = X_{T_k}$.

We again write the cover time in a clever way:
\begin{align}
\tcov &\geq \E_x[T_1 + (T_2 - T_1) + ... + 
(T_k - T_{k-1}) + ... + (T_{|A|} - T_{|A|-1})] . \nonumber
\end{align}

Our goal is to bound the cover time in terms of $\tmina$.

Consider the first term of the sum.
With probability $1/|A|$, $\sigma_1 = x$ and $T_1=0$.
Otherwise, the expected time from $x$ to $T_1$ is less than or equal to
$\tmina$.
Then
\begin{align}
\E_x[T_1] \geq \frac{1}{|A|}0 + \frac{|A| -1}{|A|} \tmina
= \left(1 - \frac{1}{|A|} \right) \tmina \nonumber
\end{align}

Now consider the $k^{th}$ term of the sum for $k > 1$.
By conditional expectation,
\begin{align}
\E_x[T_k - T_{k-1}] &=
\sum_{i=1}^k \E_x[T_k - T_{k-1} | L(k) = \sigma_i]
P(L_k = \sigma_i). \nonumber 
\end{align}

Assume that the last vertex $L_k$ to be visted among the first
$k$ vertices is not $\sigma_k$.
Then $\E_x[T_k-T_{k-1}|L(k) \neq \sigma_k] = 0$.
It follows that
\begin{align}
\E_x[T_k - T_{k-1}] &= 0 +
\E_x[T_k - T_{k-1} | L(k) = \sigma_k]
P(L_k = \sigma_k). \nonumber 
\end{align}

Since $\sigma$ is a random permutation, $P(L_k = \sigma_k) = 1/k$.

Now find the expected time between $T_{k-1}$ and $T_{k}$
given that last vertex we visit is $\sigma_k$.
Consider $r, s \in A$ such that $L_{k-1} = r$ and $L_k = s$.
Then $\E_x[T_k - T_{k-1} | L(k) = \sigma_k] = \E_r[T_s] \geq \tmina$.
So
\begin{align}
\E_x[T_k - T_{k-1}] &\geq
\tmina \frac{1}{k}. \nonumber 
\end{align}

Putting it all together,
\begin{align}
\tcov &\geq \tmina \left( 1 - \frac{1}{|A|}+ \frac{1}{2} + ... + \frac{1}{k} 
+ ... + \frac{1}{|A|-1} + \frac{1}{|A|}\right). \nonumber \\
&= \tmina \left( 1 + \frac{1}{2} + ... + \frac{1}{k} 
+ ... + \frac{1}{|A|-1} \right). \nonumber
\end{align}
\end{proof}
